---
title: "Homework 1"
subtitle: "Air Quality Dataset Analysis"

author: "Gary Walsh - 19210998"
format: pdf
header-includes:
  - \usepackage{xcolor}
  - \definecolor{darkblue}{RGB}{0, 0, 139}
  - \definecolor{darkred}{RGB}{140, 0,0}
  - \usepackage{colortbl}
  - \usepackage{booktabs}
---

# \textcolor{darkred} {Introduction}

The `airquality` dataset from the `datasets` package in R contains daily air quality measurements in New York, from May to September 1973. The dataset includes measurements of Ozone, Solar Radiation, Wind, and Temperature.

In this document, we will plot the relationship between Ozone levels and Wind speed. The data will be color-coded based on the month to provide additional insights.

## \textcolor{darkred} {Approach}

1.  **Dataset:** We use the `airquality` dataset from the `datasets` package.
2.  **Plot Type:** Scatter plot with color-coding for months and a smoother line to show the trend.
3.  **Descriptive Statistics:** Mean Ozone levels and mean Wind speed.

## \textcolor{darkred} {Descriptive Statistics}

Table 1 below presents some descriptive statistics for the airquality dataset.

```{r}
#| echo: false
#| warning: false
library(knitr)

# Calculate descriptive statistics
descriptive_stats <- data.frame(
  Statistic = c("Mean", "Median", "Std Dev", "Min", "Max"),
  Ozone = c(mean(airquality$Ozone, na.rm = TRUE), 
            median(airquality$Ozone, na.rm = TRUE), 
            sd(airquality$Ozone, na.rm = TRUE), 
            min(airquality$Ozone, na.rm = TRUE), 
            max(airquality$Ozone, na.rm = TRUE)),
  Wind = c(mean(airquality$Wind, na.rm = TRUE), 
           median(airquality$Wind, na.rm = TRUE), 
           sd(airquality$Wind, na.rm = TRUE), 
           min(airquality$Wind, na.rm = TRUE), 
           max(airquality$Wind, na.rm = TRUE))
)

# Display the table
kable(descriptive_stats, caption = "Descriptive Statistics for Ozone and Wind")
```

### \textcolor{darkblue} {Summary}

### \textcolor{darkblue}{
Ozone levels exhibit high variability with extreme values, suggesting significant fluctuations in air quality. Wind speeds are more stable, with measurements closely clustering around the mean.}

## \textcolor{darkred} {Monthly Summary Statistics for Ozone Levels}

Table 2 below presents the monthly summary statistics for Ozone levels.

```{r}
#| echo: false
#| warning: false
library(kableExtra)

# Calculate monthly summary statistics
monthly_stats <- aggregate(Ozone ~ Month, data = airquality, function(x) {
  c(Mean = mean(x, na.rm = TRUE), 
    Median = median(x, na.rm = TRUE), 
    `Std Dev` = sd(x, na.rm = TRUE), 
    Min = min(x, na.rm = TRUE), 
    Max = max(x, na.rm = TRUE))
})

# Reshape the data for better presentation
monthly_stats <- do.call(data.frame, monthly_stats)
names(monthly_stats) <- c("Month", "Mean", "Median", "Std_Dev", "Min", "Max")

# Display the table
kable(monthly_stats, caption = "Monthly Summary Statistics for Ozone Levels", format = "markdown")

```



