---
title: "Homework 1"
subtitle: "Air Quality Dataset Analysis"

author: "Gary Walsh - 19210998"
format: pdf
header-includes:
  - \usepackage{xcolor}
  - \definecolor{darkblue}{RGB}{0, 0, 139}
  - \definecolor{darkred}{RGB}{140, 0,0}
  - \usepackage{colortbl}
  - \usepackage{booktabs}
---

# \textcolor{darkred} {Introduction}

The `airquality` dataset from the `datasets` package in R contains daily air quality measurements in New York, from May to September 1973. The dataset includes measurements of Ozone, Solar Radiation, Wind, and Temperature.

In this document, we will plot the relationship between Ozone levels and Wind speed. The data will be color-coded based on the month to provide additional insights.

## \textcolor{darkred} {Approach}

1.  **Dataset:** We use the `airquality` dataset from the `datasets` package.
2.  **Plot Type:** Scatter plot with color-coding for months and a smoother line to show the trend.
3.  **Descriptive Statistics:** Mean Ozone levels and mean Wind speed.

## \textcolor{darkred} {Descriptive Statistics}

Table 1 below presents some descriptive statistics for the airquality dataset.

```{r}
#| echo: false
#| warning: false
library(knitr)

# Calculate descriptive statistics
descriptive_stats <- data.frame(
  Statistic = c("Mean", "Median", "Std Dev", "Min", "Max"),
  Ozone = c(mean(airquality$Ozone, na.rm = TRUE), 
            median(airquality$Ozone, na.rm = TRUE), 
            sd(airquality$Ozone, na.rm = TRUE), 
            min(airquality$Ozone, na.rm = TRUE), 
            max(airquality$Ozone, na.rm = TRUE)),
  Wind = c(mean(airquality$Wind, na.rm = TRUE), 
           median(airquality$Wind, na.rm = TRUE), 
           sd(airquality$Wind, na.rm = TRUE), 
           min(airquality$Wind, na.rm = TRUE), 
           max(airquality$Wind, na.rm = TRUE))
)

# Display the table
kable(descriptive_stats, caption = "Descriptive Statistics for Ozone and Wind")
```

### \textcolor{darkblue} {Summary}

### \textcolor{darkblue}{
Ozone levels exhibit high variability with extreme values, suggesting significant fluctuations in air quality. Wind speeds are more stable, with measurements closely clustering around the mean.}

## \textcolor{darkred} {Plot 1}

The scatter plot in fig 1 below shows the relationship between Ozone levels and Wind speed. The points are color-coded based on the month, and a smoother line is added to illustrate the trend.

```{r}
#| echo: false
#| warning: false
# Load the dataset
data(airquality)

# Convert Month to a factor for better plotting
airquality$Month <- factor(airquality$Month, 
                           levels = 5:9, 
                           labels = c("May", "June", "July", "August", "September"))

# Create the plot
library(ggplot2)
ggplot(airquality, aes(x = Wind, y = Ozone, color = Month)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linetype = "dashed") +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Ozone Levels vs Wind Speed",
       subtitle = "Color-coded by Month",
       x = "Wind Speed (mph)",
       y = "Ozone Levels (ppb)",
       color = "Month") +
  theme_minimal() +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        axis.title = element_text(face = "bold"),
        panel.grid.major = element_line(color = "grey80"),
        panel.grid.minor = element_line(color = "grey90"))

```

### \textcolor{darkblue} {Interpretation of the Plot}

\textcolor{darkblue} {1. Inverse Relationship:} \textcolor{darkblue}{Ozone levels generally decrease as Wind speed increases, suggesting higher wind speeds help disperse Ozone.}

\textcolor{darkblue} {2. Seasonal Variation:}

-   \textcolor{darkblue}{ May and June: Lower Ozone levels across various wind speeds.}
-   \textcolor{darkblue}{July and August: Higher Ozone levels at lower wind speeds, indicating severe Ozone pollution in mid-summer.}
-   \textcolor{darkblue} {September: Mixed data but overall lower Ozone levels as fall approaches.}

\textcolor{darkblue} {2. Data Dispersion:}

-   \textcolor{darkblue}{High variability in Ozone levels at lower wind speeds, especially in July and August.}
-   \textcolor{darkblue}{Reduced variability and lower Ozone levels at higher wind speeds.}

## \textcolor{darkred} {Monthly Summary Statistics for Ozone Levels}

Table 2 below presents the monthly summary statistics for Ozone levels.

```{r}
#| echo: false
#| warning: false
library(kableExtra)

# Calculate monthly summary statistics
monthly_stats <- aggregate(Ozone ~ Month, data = airquality, function(x) {
  c(Mean = mean(x, na.rm = TRUE), 
    Median = median(x, na.rm = TRUE), 
    `Std Dev` = sd(x, na.rm = TRUE), 
    Min = min(x, na.rm = TRUE), 
    Max = max(x, na.rm = TRUE))
})

# Reshape the data for better presentation
monthly_stats <- do.call(data.frame, monthly_stats)
names(monthly_stats) <- c("Month", "Mean", "Median", "Std_Dev", "Min", "Max")

# Display the table
kable(monthly_stats, caption = "Monthly Summary Statistics for Ozone Levels", format = "markdown")

```

## \textcolor{darkred} {Additional Simple Plot: Ozone Levels Distribution}

The histogram in fig 3 below shows the distribution of Ozone levels.

```{r}
#| echo: false
#| warning: false
ggplot(airquality, aes(x = Ozone)) +
  geom_histogram(binwidth = 10, fill = "darkblue", color = "white", alpha = 0.7) +
  labs(title = "Distribution of Ozone Levels",
       x = "Ozone Levels (ppb)",
       y = "Frequency") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        axis.title = element_text(face = "bold"))
```
